Computing exact restricted isometry constants via\\ mixed integer semidefinite programming

### 2016-03-10 RIP-MISDP-Paper/IEEE_RIP-MISDP.tex
removed sentence about presenting a methodology for MISDP
further changes in the order of d'Aspremont 08 and 07
small corrections

The goal of this paper is to compute optimal restricted isometry constants
via a mixed integer semidefinite program (MISDP). To this end, we will
introduce an MISDP with a rank-1 constraint. In order to
utilize this MISDP, we will show that the rank constraint can be relaxed
without changing the optimal value. Moreover, we discuss a solution
methodology to solve the resulting MISDP. It will turn out that
the resulting problems are quite hard to solve, even for matrices of
smaller size. However, the results allow an evaluation of the SDP
relaxations of d'Aspremont et al. Depending on the input matrices, these
bounds can be quite tight or weak. This complements the theoretical
approximation guarantees developed by d'Aspremont et
al. in~\cite{AspBG14}.

In the following, we will start by introducing the restricted isometry
property and corresponding results about the equivalence of~\eqref{l0min}
and~\eqref{l1min} in Section~\ref{sec:RIP}. In Section~\ref{sec:SDPrelax},
we will then review the SDP-relaxations of~\cite{Asp07}
and~\cite{Asp08}. Afterwards we will construct a mixed integer semidefinite
programm and show that it produces exact restricted isometry constants in
Section~\ref{sec:MISDP}. Finally, in Section~\ref{sec:numerical_results},
we present numerical results and comparisons with the SDP relaxations in
\cite{Asp07} and \cite{Asp08}.

@@

As one way to solve (\ref{phi}) again the semidefinite lifting technique is used in \cite{Asp08} to compute an upper bound on $\phi$ via
\begin{align}\label{Asp08}
 \max \quad & \sum_{i=1}^n\Tr(P_iB_i) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & X \succeq 0 \tag{A2-Primal} \\
 & X \succeq P_i \succeq 0 \quad \text{for } i = 1, ..., n, \nonumber
\end{align}
for $B_i = b_ib_i\T - \rho I,$ where $b_i$ is the $i$-th column of a squareroot of $\Sigma$. The corresponding dual problem can then be written as 
\begin{align}\label{Asp08Dual}
  \min \quad & \lambda_{\max}\left(\sum_{i=1}^n Y_i \right) \nonumber \\
  \text{s.t.} \quad & Y_i \succeq B_i \quad \forall i \leq n \tag{A2-Dual} \\
  & Y_i \succeq 0 \quad \ \ \text{for } i = 1, ..., n, \nonumber
\end{align}
which only has half as many variables as (\ref{Asp08}) and one less linear and semidefinite constraint. Then (\ref{Asp08Dual}) can be solved for non-negative $\rho$, which should be smaller than $\Sigma_{11}$, because otherwise the
optimal solution for (\ref{phi}) will always be $x=0$.

@@

In the following we will to show that (\ref{MISDP}) not only computes a lower bound but the exact value of $\alpha_k$. For this we will use results about the geometry of semidefinite programs to show that there always exists an 
optimal solution to (\ref{MISDP}) with rank 1 so that the optimal objective values of (\ref{Rk1MISDP}) and (\ref{MISDP}) have to agree. We will use the following result by Pataki \cite{pat98} about the relationship between the rank of
solutions of semidefinite programs and the dimension of corresponding faces, meaning a subset $F$ of the convex set $P$ such that \mbox{$x \in F, y, z \in P$} and $x = 0.5y + 0.5z$ imply $y,z \in F$:

@@

\begin{theorem}\label{Rk1thm}
 For every $A \in \R^{m \times n}$ and $k \geq 1$ there exists an optimal solution $X^*, z^*$ to (\ref{MISDP}) such that $\Rk(X^*) = 1$. 
\end{theorem}
\begin{proof}
 Let $\hat{X}, \hat{z}$ be an optimal solution to (\ref{MISDP}) with $\sum_{i=1}^n z_i = \ell \leq k$ 
 (because of $\Tr(X) = 1$ the problem is bounded and $X_{11} = 1, X_{ij} = 0 \ \forall i,j \neq 1,$ \mbox{$z_1 = 1, z_i = 0 \ \forall i \neq 1$} is always feasible for $k > 0$, so an optimal solution exists).
 Define
 \begin{align}\label{Proj}
 \min \quad & \Tr(\tilde{A}\T\tilde{A}\tilde{X}) \nonumber \\
 \text{s.t.} \quad & \Tr(\tilde{X}) = 1 \tag{Proj} \\
 & \tilde{X} \succeq 0 , \tilde{X} \in S_\ell, \nonumber 
 \end{align}\comment{$\tilde{X} \in S_\ell$ eigentlich in $\tilde{X} \succeq 0$ enthalten, aber betont halt Dimension, so drinnlassen oder \"andern?}
where $\tilde{A} = (A_{i_1}, A_{i_2}, ..., A_{i_\ell})$ consists of those
columns $\{{i_1}, {i_2}, ..., {i_\ell}\} \enifed T$ of $A$ for which
$\hat{z}_{i_k} = 1$.

@@

So the optimal objective values of (\ref{Proj}) and (\ref{MISDP}) agree and with this construction we also get that 
$\Rk(X^{\prime}) = \Rk(\check{X})$, so it suffices to show that (\ref{Proj}) always allows for a solution of rank one.
Let $\breve{X}$ be an extreme point of (\ref{Proj}) (as the set of optimal points of (\ref{Proj}) is a non-empty, convex, compact subset of $\R^{\ell(\ell+1)/2}$, such a point exists
\cite[Corollary 18.5.1]{Roc70}) with corresponding face $F=\{\breve{X}\}$. Then using Theorem~\ref{patakiLemma} on (\ref{Proj}) with $\tilde{m} = 1$ LP-constraint gives
\begin{equation*}
 \frac{1}{2}\Rk(\breve{X})\cdot(\Rk(\breve{X})+1) \leq \tilde{m} + \textbf{dim}(F) = 1 + 0 = 1.
\end{equation*}
Since $\breve{X}=0$ is infeasible for $\Tr(X)=1$, we must have $\Rk(\breve{X}) = 1$, which can then be lifted to an optimal solution of (\ref{MISDP}) with rank one.
\end{proof}

@@

\begin{remark}\label{HalfConstraint}
We can substitute the constraint $-z_j \leq X_{ij} \leq z_j$ for all non-diagonal entries of $X$ by the stronger constraint 
\begin{equation}\label{eq:HalfConstraint}
  -\frac{1}{2}z_j \leq X_{ij} \leq \frac{1}{2}z_j.
\end{equation}
Let $X$ satisfy $\Tr(X)=1$ and $X\succeq 0$, which immediately implies $X_{ii}\geq 0$.
Since all minors of $X$ need to be nonnegative, we get that
\begin{equation*}
  X_{ij}^2 \leq  X_{ii} X_{jj} \leq \frac{1}{4}
\end{equation*}
and therefore $-1/2 \leq X_{ij} \leq 1/2$, which implies the equivalence between
$-z_j \leq X_{ij} \leq z_j$ and~\eqref{eq:HalfConstraint} for all integer values 
of $z$, while~\eqref{eq:HalfConstraint} might cut off some solutions for fractional $z$.
\end{remark}

>>>>

The goal of this paper is to compute optimal restricted isometry constants
via a mixed integer semidefinite program (MISDP). To this end, we will
introduce an MISDP with a rank-1 constraint. In order to
utilize this MISDP, we will show that the rank constraint can be relaxed
without changing the optimal value. It will turn out that
the resulting problems are quite hard to solve, even for matrices of
smaller size. However, the results allow an evaluation of the SDP
relaxations of d'Aspremont et al. Depending on the input matrices, these
bounds can be quite tight or weak. This complements the theoretical
approximation guarantees developed by d'Aspremont et
al. in~\cite{AspBG14}.

In the following, we will start by introducing the restricted isometry
property and corresponding results about the equivalence of~\eqref{l0min}
and~\eqref{l1min} in Section~\ref{sec:RIP}. In Section~\ref{sec:SDPrelax},
we will then review the SDP-relaxations of~\cite{Asp08}
and~\cite{Asp07}. Afterwards we will construct a mixed integer semidefinite
programm and show that it produces exact restricted isometry constants in
Section~\ref{sec:MISDP}. Finally, in Section~\ref{sec:numerical_results},
we present numerical results and comparisons with the SDP relaxations in
\cite{Asp08} and \cite{Asp07}.

@@

As one way to solve (\ref{phi}) the semidefinite lifting technique is used again in \cite{Asp08} to compute an upper bound on $\phi$ via
\begin{align}\label{Asp08}
 \max \quad & \sum_{i=1}^n\Tr(P_iB_i) \nonumber \\
 \text{s.t.} \quad & \Tr(X) = 1 \nonumber \\
 & X \succeq 0 \tag{A2-Primal} \\
 & X \succeq P_i \succeq 0 \quad \text{for } i = 1, ..., n, \nonumber
\end{align}
for $B_i = b_ib_i\T - \rho I,$ where $b_i$ is the $i$-th column of a squareroot of $\Sigma$. The corresponding dual problem can then be written as 
\begin{align}\label{Asp08Dual}
  \min \quad & \lambda_{\max}\left(\sum_{i=1}^n Y_i \right) \nonumber \\
  \text{s.t.} \quad & Y_i \succeq B_i \quad \forall i \leq n \tag{A2-Dual} \\
  & Y_i \succeq 0 \quad \ \ \text{for } i = 1, ..., n, \nonumber
\end{align}
which only has half as many variables as (\ref{Asp08}) and one less linear and semidefinite constraint. Then (\ref{Asp08Dual}) can be solved for non-negative $\rho$, which should be smaller than $\Sigma_{11}$, because otherwise the
optimal solution for (\ref{phi}) will always be $x=0$.

@@

In the following we will show that (\ref{MISDP}) not only computes a lower bound but the exact value of $\alpha_k$. For this we will use results about the geometry of semidefinite programs to show that there always exists an 
optimal solution to (\ref{MISDP}) with rank 1 so that the optimal objective values of (\ref{Rk1MISDP}) and (\ref{MISDP}) have to agree. We will use the following result by Pataki \cite{pat98} about the relationship between the rank of
solutions of semidefinite programs and the dimension of corresponding faces, meaning a subset $F$ of the convex set $P$ such that \mbox{$x \in F, y, z \in P$} and $x = 0.5y + 0.5z$ imply $y,z \in F$:

@@

\begin{theorem}\label{Rk1thm}
 For every $A \in \R^{m \times n}$ and $k \geq 1$ there exists an optimal solution $X^*, z^*$ to (\ref{MISDP}) such that $\Rk(X^*) = 1$. 
\end{theorem}
\begin{proof}
 Because of the constraint $\Tr(X) = 1$, Problem~\eqref{MISDP} is bounded and $X_{11} = 1, X_{ij} = 0 \ \forall i,j \neq 1,$ \mbox{$z_1 = 1, z_i = 0 \ \forall i \neq 1$} 
 is always feasible for $k > 0$,  so an optimal solution to~\eqref{MISDP} exists.
 Let $\hat{X}, \hat{z}$ be such an optimal solution to (\ref{MISDP}) with $\sum_{i=1}^n z_i =: \ell \leq k$.
 Define
 \begin{align}\label{Proj}
 \min \quad & \Tr(\tilde{A}\T\tilde{A}\tilde{X}) \nonumber \\
 \text{s.t.} \quad & \Tr(\tilde{X}) = 1 \tag{Proj} \\
 & \tilde{X} \succeq 0  \nonumber 
 \end{align}
for $\tilde{X} \in S_\ell$, where $\tilde{A} = (A_{i_1}, A_{i_2}, ..., A_{i_\ell})$ consists of those
columns $\{{i_1}, {i_2}, ..., {i_\ell}\} \enifed T$ of $A$ for which
$\hat{z}_{i_k} = 1$.

@@

So the optimal objective values of (\ref{Proj}) and (\ref{MISDP}) agree and with this construction also 
$\Rk(X^{\prime}) = \Rk(\check{X})$, so it suffices to show that (\ref{Proj}) always allows for an optimal solution of rank one.
Let $\breve{X}$ be an extreme point of (\ref{Proj}) (as the set of optimal points of (\ref{Proj}) is a non-empty, convex, compact subset of $\R^{\ell(\ell+1)/2}$, such a point exists
\cite[Corollary 18.5.1]{Roc70}) with corresponding face $F=\{\breve{X}\}$. Then using Theorem~\ref{patakiLemma} on (\ref{Proj}) with $\tilde{m} = 1$ LP-constraint gives
\begin{equation*}
 \frac{1}{2}\Rk(\breve{X})\cdot(\Rk(\breve{X})+1) \leq \tilde{m} + \textbf{dim}(F) = 1 + 0 = 1.
\end{equation*}
Since $\breve{X}=0$ is infeasible for $\Tr(X)=1$, we must have $\Rk(\breve{X}) = 1$, which can then be lifted to an optimal solution of (\ref{MISDP}) with rank one.
\end{proof}

@@

\begin{remark}\label{HalfConstraint} \comment{Als Remark belassen oder ein Lemma draus machen?}
We can substitute the constraint $-z_j \leq X_{ij} \leq z_j$ for all non-diagonal entries of $X$ by the stronger constraint 
\begin{equation}\label{eq:HalfConstraint}
  -\frac{1}{2}z_j \leq X_{ij} \leq \frac{1}{2}z_j.
\end{equation}
Let $X$ satisfy $\Tr(X)=1$ and $X\succeq 0$, which immediately implies $X_{ii}\geq 0$.
Since all minors of $X$ need to be nonnegative, we get that
\begin{equation*}
  X_{ij}^2 \leq  X_{ii} X_{jj} \leq \frac{1}{4}
\end{equation*}
and therefore $-1/2 \leq X_{ij} \leq 1/2$, which implies the equivalence between
$-z_j \leq X_{ij} \leq z_j$ and~\eqref{eq:HalfConstraint} for all integer values 
of $z$, while~\eqref{eq:HalfConstraint} might cut off some solutions for fractional $z$.
\end{remark}
